---
title: "Plotting IAMC-format MESSAGE outputs"
authour: "Jihoon Min"
date: "Feb 19, 2020"
output: html_document
---
This document is about creating some example plots of key variables of MESSAGE scenarios given in IAMC-format snapshots. This was created to prepare the webinar for the UNEP-FI project on Jan 27, 2020.

```{r include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=TRUE)  # Do not output the intermediate logs in the final html
```

# Import and rearrange data for ggplot
```{r}
library(tidyverse)
library(readxl)
library(ggplot2)

# The IAMC-format snapshot with initial scenarios used for UNEP-FI webinar. 
datapath = "C:/Users/min/IIASA/VAN RUIJVEN Bas - Finance/UNEP FI/Data/CDLINKS_stocktaking_MESSAGE_native_regions_FULL.csv"

raw_df = read_csv(datapath) %>% filter(REGION == "World") %>% 
  mutate(SSP = case_when(grepl("SSP1", SCENARIO) ~ "SSP1", 
                         grepl("SSP3", SCENARIO) ~ "SSP3",
                         TRUE ~ "SSP2")) %>%
  mutate(budget = case_when(grepl("1600", SCENARIO) ~ 1600, 
                         grepl("1000", SCENARIO) ~ 1000,
                         grepl("400", SCENARIO) ~ 400,
                         TRUE ~ Inf)) %>% 
  select(-MODEL, -REGION) %>%
  mutate(SCENARIO=sub("_V4", "", SCENARIO)) %>% # SCENARIO naems are too long. 
  mutate(SCENARIO=ifelse(grepl("SSP1|SSP3", SCENARIO), SCENARIO, paste0(SCENARIO, "_SSP2"))) %>% 
  mutate(Category = case_when(grepl("2020", SCENARIO) ~ "NPi2020",   # Group the scenarios to remove redundant information on figures.
                              grepl("2030", SCENARIO) ~ "INDC2030",
                              grepl("NPi_", SCENARIO) ~ "NPi",
                              grepl("INDCi_", SCENARIO) ~ "INDCi",
                              grepl("NoPolicy", SCENARIO) ~ "NoPolicy")) %>%
  mutate(Category = factor(Category,   # Order by types of carriers
                         levels=c("NoPolicy", "NPi", "INDCi", "NPi2020", "INDC2030"),
                         labels=c("NoPolicy", "NPi", "INDCi", "NPi2020", "INDC2030"))) 

emissions_df = raw_df %>% filter(VARIABLE %in% c("Emissions|Kyoto Gases"))  #"Emissions|CO2", "Emissions|CH4", "Emissions|N2O", 
p_carbon_df = raw_df %>% filter(VARIABLE %in% c("Price|Carbon"))

# Long format
emissions_df_l = emissions_df %>% gather(key = "year", value = "emissions", "2005":"2100", -SCENARIO, -SSP, -budget, -VARIABLE, -Category) 
p_carbon_df_l = p_carbon_df %>% gather(key = "year", value = "p_carbon", "2005":"2100", -SCENARIO, -SSP, -budget, -VARIABLE, -Category) 
```

# Plot emissions & carbon price

```{r}
em = ggplot() +
  geom_line(data = emissions_df_l, aes(x=year, y=emissions, group=SCENARIO, color = Category), size=1.3, alpha = 0.7) + 
  geom_line(data = p_carbon_df_l %>% filter(budget!=Inf),  # Carbon price line
            aes(x=year, y=p_carbon*50, group=SCENARIO, color = Category), size=1, alpha = 0.4, linetype="twodash") + 
  facet_grid(budget~SSP) +
  labs(y="CO2e emissions [Mt/yr]") + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  scale_y_continuous(sec.axis = sec_axis(~./50, name = "Carbon Price [US$2010/t CO2]"))
print(em)

# Export as PDF
pdf(file = paste0("./Plot/GHG emissions.pdf"), width = 10.46, height = 6.5)
print(em)  
dev.off()

# Export as PNG
png(file = paste0("./Plot/GHG emissions.png"), width = 1600, height = 1200)
em + theme(text = element_text(size=30))
dev.off()

```

# Arrange data for primary energy plot
```{r}
# Define plotting order of scenarios
Scenario_order = c("NPi2020_SSP1", "INDC2030i_SSP1", "NPi2020_SSP2", "INDC2030i_SSP2", "NPi2020_SSP3", "INDC2030i_SSP3", 
                                              "NoPolicy_SSP1","NPi_SSP1", "INDCi_SSP1", 
                                              "NoPolicy_SSP2","NPi_SSP2", "INDCi_SSP2",
                                              "NoPolicy_SSP3","NPi_SSP3", "INDCi_SSP3")

# Derine plotting order of energy categories                   
primary_cr = rev(c("Coal","Oil","Gas", "Nuclear", "Hydro", "Solar", "Wind", 
               "Geothermal", "Ocean", "Biomass", "Non-Biomass Renewables", "Other", "Secondary Energy Trade"))

# Prepare dataframe for ggplot
primary_df = raw_df %>% filter(grepl("Primary Energy", VARIABLE)) %>% 
  mutate(lvl = str_count(VARIABLE, "\\|"), VARIABLE=sub("Primary Energy\\|", "", VARIABLE)) %>%  # Shorten the VARIABLE names
  filter(lvl==1) %>%  # Select only the first level category of PE
  filter(!grepl("Fossil|Trade", VARIABLE)) %>% # Remove "Primary Energy|Fossil" because of double counting, also elec trade to avoid unnecessary confusion
  select(-lvl) %>%
  mutate(VARIABLE=factor(VARIABLE,   # Order by types of carriers
                         levels=primary_cr,
                         labels=primary_cr)) %>%
  select(-c("2005", "2010", "2060", "2080", "2090")) # Remove unnecessary years

# Converty to long format
primary_df_l = primary_df %>% gather(key = "year", value = "EJ", "2020":"2100", -SCENARIO, -SSP, -budget, -VARIABLE) %>%
  mutate(SCENARIO = gsub("_400|_1000|_1600", "", SCENARIO)) %>%
  mutate(SCENARIO = factor(SCENARIO, levels=Scenario_order, labels=Scenario_order))  # Reorder scenario labels for better display

# Have full combinations of SSP&budget, so that we show which scenarios are infeasible on the plots (as empty spaces)
full_comb1 = list(SCENARIO=Scenario_order[1:6], budget=c(400, 1000, 1600), year=unique(primary_df_l$year), VARIABLE=unique(primary_df_l$VARIABLE))
full_comb2 = list(SCENARIO=Scenario_order[7:15], budget=Inf, year=unique(primary_df_l$year), VARIABLE=unique(primary_df_l$VARIABLE))
full_comb = rbind(do.call(expand.grid, full_comb1), do.call(expand.grid, full_comb2))
primary_df_l = full_comb %>% left_join(primary_df_l)

# Separate the dataframe by carbon budget level to make them as separate graphs
pe_list = lapply(unique(primary_df_l$budget), function(b) {primary_df_l %>% filter(budget==b)})
names(pe_list) = as.character(unique(primary_df_l$budget))

```

# Plot primary energy by energy carrier & scenario & budget

```{r}
#Define color schemes for primary energy source
library(RColorBrewer)
getPalette = colorRampPalette(brewer.pal(11, "Spectral"))
CompColors <- rev(getPalette(length(primary_cr)))
names(CompColors) <- primary_cr
CompColScale <- scale_fill_manual(values = CompColors)

# Plot and export
for (i in 1:length(pe_list)) {
  pe = ggplot(pe_list[[i]], aes(x=SCENARIO, y=EJ, fill=VARIABLE)) +
    geom_bar(stat = 'identity', position = 'stack') + 
    facet_grid(~year) +
    labs(y="Primary Energy [EJ/yr]", title = paste(names(pe_list)[i], "GtCO₂ budget")) +
    lims(y=c(-100, 2000)) + 
    CompColScale +
    theme(axis.text.x = element_text(angle = 60, hjust = 1))
  print(pe)  
  
  pdf(file = paste0("./Plot/Primary energy ", names(pe_list)[i], ".pdf"), width = 10.46, height = 6.5)
  print(pe)
  dev.off()
  
  png(file = paste0("./Plot/Primary energy ", names(pe_list)[i], ".png"), width = 1900, height = 1200)
  print(pe + theme(text = element_text(size=30)))
  dev.off()
}
```

# Arrange data for final energy plot
```{r}
# Ordering of energy carrier
final_cr = rev(c("Solids","Liquids","Gases", "Electricity", "Heat", "Solar", "Geothermal", "Hydrogen"))

# Main dataframe for final energy plot
final_df = raw_df %>% filter(grepl("Final Energy", VARIABLE)) %>% mutate(lvl = str_count(VARIABLE, "\\|")) %>% 
  filter(lvl>0) %>% # Select only the first level category of PE
  filter(!grepl("Trade|Solids\\|", VARIABLE)) %>% # Remove Electricity Trade
  mutate(VARIABLE=sub("Final Energy\\|", "", VARIABLE)) %>%
  # select(-lvl) %>%
  select(-c("2005", "2010", "2060", "2080", "2090")) # Remove unnecessary years

# Define sub-sectors for final energy use
enduse = c("Industry","Non-Energy Use", "Residential and Commercial", "Transportation")

# Function generating long-format table for each sub-sector
eu_l = function(sect) {
  df = final_df %>% # By End-use sector
    filter(grepl(sect, VARIABLE)) %>% 
    filter(lvl==2) %>%
    gather(key = "year", value = "EJ", "2020":"2100", -SCENARIO, -SSP, -budget, -VARIABLE) %>%
    mutate(SCENARIO = gsub("_400|_1000|_1600", "", SCENARIO)) %>%
    mutate(SCENARIO = factor(SCENARIO, levels=Scenario_order, labels=Scenario_order)) 
  return(df)
}

# Convert to long format by sub-sector
final_df_l_eu_list = lapply(enduse, FUN=eu_l)
names(final_df_l_eu_list) = enduse

final_df_l_cr = final_df %>% # By Carrier
  filter(!grepl("Industry|Non-Energy Use|Residential|Transportation", VARIABLE)) %>% 
  gather(key = "year", value = "EJ", "2020":"2100", -SCENARIO, -SSP, -budget, -VARIABLE) %>%
  mutate(SCENARIO = gsub("_400|_1000|_1600", "", SCENARIO)) %>%
  mutate(SCENARIO = factor(SCENARIO, levels=Scenario_order, labels=Scenario_order)) %>% # Reorder scenario labels for better display
  mutate(VARIABLE=factor(VARIABLE,   # Order by types of carriers
                         levels=final_cr,
                         labels=final_cr))

# Have full combinations of SSP/budget to show which scenarios are infeasible on the plots
full_comb1 = list(SCENARIO=Scenario_order[1:6], budget=c(400, 1000, 1600), year=unique(final_df_l_cr$year),
                  VARIABLE=unique(final_df_l_cr$VARIABLE))
full_comb2 = list(SCENARIO=Scenario_order[7:15], budget=Inf, year=unique(final_df_l_cr$year), VARIABLE=unique(final_df_l_cr$VARIABLE))
full_comb = rbind(do.call(expand.grid, full_comb1), do.call(expand.grid, full_comb2))
final_df_l_cr = full_comb %>% left_join(final_df_l_cr)

# Separate the dataframe by carbon budget level to make them as separate graphs
fe_list = lapply(unique(final_df_l_cr$budget), function(b) {final_df_l_cr %>% filter(budget==b)})
names(fe_list) = as.character(unique(final_df_l_cr$budget))
```

# Plot total final energy by energy carrier & scenario & budget
```{r}
# Define color schemes for final energy use
library(RColorBrewer)
getPalette = colorRampPalette(brewer.pal(11, "Spectral"))
CompColors_f <- rev(getPalette(length(final_cr)))
names(CompColors_f) <- final_cr

CompColScale_f <- scale_fill_manual(values = CompColors_f)

# Plot and export by budget level
for (i in 1:length(fe_list)) {
  fe = ggplot(fe_list[[i]], aes(x=SCENARIO, y=EJ, fill=VARIABLE)) +
    geom_bar(stat = 'identity', position = 'stack') + 
    facet_grid(~year) +
    labs(y="Final Energy [EJ/yr]", title = paste(names(fe_list)[i], "GtCO₂ budget")) +
    lims(y=c(0, 1300)) +
    CompColScale_f +
    theme(axis.text.x = element_text(angle = 60, hjust = 1))
  print(fe)  
  
  pdf(file = paste0("./Plot/Final energy ", names(fe_list)[i], ".pdf"), width = 10.46, height = 6.5)
  print(fe)  
  dev.off()
  
  png(file = paste0("./Plot/Final energy ", names(fe_list)[i], ".png"), width = 1900, height = 1200)
  print(fe + theme(text = element_text(size=30)))
  dev.off()
}
```

# Generate similar FE plots for each sub-sector
```{r}
for (j in 1:length(final_df_l_eu_list)) {

  df_l = final_df_l_eu_list[[j]]
  
  full_comb1 = list(SCENARIO=Scenario_order[1:6], budget=c(400, 1000, 1600), year=unique(df_l$year), 
                    VARIABLE=unique(df_l$VARIABLE))
  full_comb2 = list(SCENARIO=Scenario_order[7:15], budget=Inf, year=unique(df_l$year), 
                    VARIABLE=unique(df_l$VARIABLE))
  full_comb = rbind(do.call(expand.grid, full_comb1), do.call(expand.grid, full_comb2))
  df_l = full_comb %>% left_join(df_l)

  fe_list = lapply(unique(df_l$budget), function(b) {df_l %>% filter(budget==b)})
  names(fe_list) = as.character(unique(df_l$budget))
  
  for (i in 1:length(fe_list)) {
    fe = ggplot(fe_list[[i]], aes(x=SCENARIO, y=EJ, fill=VARIABLE)) +
      geom_bar(stat = 'identity', position = 'stack') + 
      facet_grid(~year) +
      labs(y="Final Energy [EJ/yr]", title = paste(names(fe_list)[i], "GtCO₂ budget")) +
      lims(y=c(0, 500)) +
      # CompColScale_f +
      theme(axis.text.x = element_text(angle = 60, hjust = 1))
    print(fe)  
    
    pdf(file = paste("./Plot/Final energy", names(fe_list)[i], names(final_df_l_eu_list)[j], ".pdf"), width = 10.46, height = 6.5)
    print(fe)  
    dev.off()
    
    png(file = paste("./Plot/Final energy ", names(fe_list)[i], names(final_df_l_eu_list)[j], ".png"), width = 1900, height = 1200)
    print(fe + theme(text = element_text(size=30)))
    dev.off()
  }
}
```


# Plot GDP and population assumptions for each SSP and carbon budget
```{r}
# Unit: Bil$.GDP & Mil.pop
GDP_pop_df = raw_df %>% filter(VARIABLE %in% c("Population", "GDP|PPP")) 

GDP_pop_df_l = GDP_pop_df %>% gather(key = "year", value = "val", "2005":"2100", -SCENARIO, -SSP, -budget,
                                     -VARIABLE, -Category) %>% select(-UNIT) %>% spread(VARIABLE, val) %>%
  mutate(GDP_pcap = `GDP|PPP`*1e3/Population) %>% select(-`GDP|PPP`) %>%
  mutate(SCENARIO = gsub("_400|_1000|_1600", "", SCENARIO)) %>%
  mutate(Population = Population/1000) %>% # Convert to Billion
  filter(year!=2005) # GDP@2005 = 0
```

```{r}
gp = ggplot() +
  geom_line(data = GDP_pop_df_l, aes(x=year, y=GDP_pcap, group=SCENARIO, color = Category), size=1.3, alpha = 0.7) + 
  geom_line(data = GDP_pop_df_l, aes(x=year, y=Population*10*1000, group=SCENARIO, color = Category), size=1, alpha = 0.4, linetype="twodash") + 
  facet_grid(budget~SSP) +
  labs(y="GDP per capita [US$2010 PPP/yr]") + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  scale_y_continuous(sec.axis = sec_axis(~./10/1000, name = "Population [Billion]"), limits = c(0,1.5e5)) 
print(gp)

pdf(file = paste0("./Plot/GDP-Pop.pdf"), width = 10.46, height = 6.5)
print(gp)  
dev.off()

png(file = paste0("./Plot/GDP-Pop.png"), width = 1600, height = 1200)
gp + theme(text = element_text(size=30))
dev.off()
```

 